Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-01 rnni_convexity.tex
change terminology shortest path that is caterpillar path

In this section we analyse shortest paths in $\rnni$ by using the algorithms introduced in the previous section.
They enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
In Section~\ref{section:diameter} we will investigate diameter and radius of the $\rnni$ graph.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} $rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$.
After proving this, we are going to consider the so-called Split Theorem, which was raised in \autocite{Gavryushkin2018-ol}.
We will give a counterexample to disprove their version of the Split Theorem and claim an alternative version, the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}).

@@

\begin{proof}
We prove this theorem by backwards induction on the length of a shortest caterpillar path $d_c(T,R) = l$ ($l$ is the induction parameter) between trees $T$ and $R$.
We saw in the proof of Lemma~\ref{lemma:caterpillar_dist=diameter} that the maximum possible value for $l$ is $\frac{(n-1)(n-2)}{2}$, which forms the induction basis.
\todo{Lena: Make sure we ALWAYS use this terminology}
Lemma~\ref{lemma:caterpillar_dist=diameter} also implies in this case that $d(T, R) = d_c(T, R)$, hence there is a caterpillar path between $T$ and $R$ of length $d(T, R)$.

@@

\begin{proof}
Let $T = [1,2,3,\ldots,n]$ and $R = [n-1,n,n-2,n-3, \ldots, 1]$ be two caterpillar trees.
According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
When running this algorithm to compute a path from $T$ to $R$, each taxon $i$ ($i = n, \ldots, 3$) swaps $i-2$ times with its upper neighbour.
Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=3}^{n}(i-2)  = \frac{(n-1)(n-2)}{2}$.
\end{proof}

>>>>

In this section we analyse shortest paths in $\rnni$ by using the algorithms introduced in the previous section.
They enable us to prove that between caterpillar trees $T$ and $R$ there is always a caterpillar path of length $d(T,R)$ (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
In Section~\ref{section:diameter} we will investigate diameter and radius of the $\rnni$ graph.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} $rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$.
After proving this, we are going to consider the so-called Split Theorem, which was raised in \autocite{Gavryushkin2018-ol}.
We will give a counterexample to disprove their version of the Split Theorem and claim an alternative version, the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}).

@@

\begin{proof}
We prove this theorem by backwards induction on the length of a shortest caterpillar path $d_c(T,R) = l$ ($l$ is the induction parameter) between trees $T$ and $R$.
We saw in the proof of Lemma~\ref{lemma:caterpillar_dist=diameter} that the maximum possible value for $l$ is $\frac{(n-1)(n-2)}{2}$, which forms the induction basis.
Lemma~\ref{lemma:caterpillar_dist=diameter} also implies in this case that $d(T, R) = d_c(T, R)$, hence there is a caterpillar path between $T$ and $R$ of length $d(T, R)$.

@@

\begin{proof}
Let $T = [1,2,3,\ldots,n]$ and $R = [n-1,n,n-2,n-3, \ldots, 1]$ be two caterpillar trees.
According to Theorem~\ref{thm:caterpillar_convex}, there is a caterpillar path of length $d(T,R)$ between $T$ and $R$.
We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
When running this algorithm to compute a path from $T$ to $R$, each taxon $i$ ($i = n, \ldots, 3$) swaps $i-2$ times with its upper neighbour.
Therefore, the length of a shortest caterpillar path, and therefore the distance between $T$ and $R$, is $\sum\limits_{i=3}^{n}(i-2)  = \frac{(n-1)(n-2)}{2}$.
\end{proof}
